---
title: "Cookbook for rvec Package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cookbook for rvec Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

This cookbook collects recipes for common tasks with rvecs. Each recipe
starts with a problem, gives a solution, and explains when and how to use it.

```{r libs}
library(rvec)
```

# Summarise elements within draws

## Calculate a minimum, maximum, or range {#min-max-range}

### Problem

You have draws for rates in several regions. You want draws for the lowest
rate, highest rate, or range of rates across the regions.

### Solution

Use `min()`, `max()`, or `range()`. Each function calculates its result
independently within each draw.

Here each row is a region and each column is a draw. The region with the
lowest or highest rate can differ between draws.

```{r regional-rates}
m <- rbind(
  north = c(3, 1, 8, 4),
  south = c(1, 7, 3, 6),
  west = c(2, 4, 5, 2)
)
m
rate <- rvec(m)
```

```{r extrema}
lowest <- min(rate)
highest <- max(rate)
limits <- range(rate)

as.matrix(lowest)
as.matrix(highest)
as.matrix(limits)
```

`lowest` and `highest` are rvecs of length one, each retaining all four draws.
`limits` is an rvec of length two: its first element contains the minima and
its second contains the maxima. `range()` returns these two endpoints;
use `max(rate) - min(rate)` for the distance between them.

You can then summarise the draws of a derived quantity, just as with any
other rvec:

```{r summarise-highest}
draws_mean(highest)
```

### Discussion

`min(rate)` and `draws_min(rate)` answer different questions. `min(rate)`
finds the lowest regional rate within each draw. `draws_min(rate)` finds
the lowest draw for each region, returning an ordinary vector with one
value per region.

```{r compare-directions}
draws_min(rate)
```

Use `na.rm = TRUE` to remove missing values before calculating extrema.
For `range()`, `finite = TRUE` excludes missing values and both infinities.

You can supply additional rvecs or ordinary vectors, but put an rvec first.
An ordinary value is included in every draw:

```{r include-reference}
min(rate, 0)
```

### See Also

- `?min`, `?max`, and `?range` for base R's treatment of missing and empty inputs.
- `?draws_min` for minima and maxima across draws.
- [Troubleshooting: put the rvec first in `min()`](troubleshooting.html#min-argument-order).


## Find the position of an extreme value {#extreme-positions}

### Problem

You want to know which region has the largest rate in each draw. The region
may differ between draws.

### Solution

Use `which.max()` for the position of the first maximum in each draw, or
`which.min()` for the first minimum.

```{r extreme-positions}
rate <- rvec(rbind(
  north = c(3, 1, 8, 4),
  south = c(1, 7, 3, 6),
  west = c(2, 4, 5, 2)
))
as.matrix(which.max(rate))
as.matrix(which.min(rate))
```

The result is a length-one rvec. Its values are positions in the original
element order: `1` is north, `2` is south, and `3` is west in this example.

### Discussion

Missing values are ignored. If a nonempty draw has no valid value, its index
is `NA` and a warning reports how many draws were affected. A zero-element
rvec returns a zero-length rvec without a warning. Ties select the first
position.

The index may differ between draws, so it cannot generally serve as one
ordinary row subscript for a data frame.

### See Also

- `?which.min` for missing values, ties, and character coercion.
- [Calculate a minimum, maximum, or range](#min-max-range).


## Calculate quantiles within each draw {#quantiles-within-draws}

### Problem

You have draws for rates in several regions and want draws for the median
and quartiles across those regions.

### Solution

Use `quantile()`, specifying the probabilities you need. Each column below
is a draw, and each row is a region.

```{r quantile-input}
rate <- rvec(rbind(
  north = c(3, 1, 8, 4),
  south = c(1, 7, 3, 6),
  west = c(2, 4, 5, 2)
))
regional_quantiles <- quantile(rate, probs = c(0.25, 0.5, 0.75))
as.matrix(regional_quantiles)
```

The result is an rvec of length three, with one element for each probability
and all four draws retained. Probability labels supply its names; use
`names = FALSE` to omit them.

### Discussion

`quantile(rate)` summarises regions within each draw. In contrast,
`draws_quantile(rate)` summarises draws separately for each region and returns
a tibble:

```{r quantile-directions}
draws_quantile(rate, probs = c(0.25, 0.5, 0.75))
```

Use `na.rm = TRUE` to remove missing values within each draw. The `type`
argument selects among base R's nine quantile algorithms; the default is
`type = 7`. For example, `type = 1` selects observed values rather than
interpolating between them.

```{r quantile-algorithm}
as.matrix(quantile(rate, probs = c(0.25, 0.5, 0.75), type = 1))
```

### See Also

- `?quantile` for probabilities, algorithms, and missing-value handling.
- `?draws_quantile` for quantiles across draws.
- [Calculate a minimum, maximum, or range](#min-max-range).


# Modify elements within draws

## Impose lower and upper bounds {#bound-values}

### Problem

You want to impose a bound on every value in an rvec, retaining all elements
and draws.

### Solution

Use `pmax()` for a lower bound and `pmin()` for an upper bound. Combine them
to impose both bounds:

```{r bound-values}
x <- rvec(rbind(a = c(-2, 0.3), b = c(4, -1)))
as.matrix(pmax(x, 0))
as.matrix(pmin(x, 1))
as.matrix(pmin(pmax(x, 0), 1))
```

An rvec can occur in any argument position, so `pmax(0, x)` also works.
You can supply a separate bound for each element:

```{r element-bounds}
as.matrix(pmax(x, c(0, 0.5)))
```

### Discussion

`pmax(x, 0)` compares each value with zero. In contrast, `max(x, 0)`
summarises all elements and zero within each draw, returning a length-one
rvec.

With rvec arguments, element lengths must agree or be one, and draw counts
must agree or be one. Ordinary vectors supply values for elements and are
repeated across draws. Names follow the first argument, so swapping arguments
may change names even when the values are unchanged.

Missing values are retained by default. Use `na.rm = TRUE` to ignore a
missing input when another corresponding value is available; if all are
missing, the result remains missing.

The rvec package supplies wrappers for these functions. Calls without rvecs
retain base R behaviour; explicitly calling `base::pmin()` or `base::pmax()`
bypasses rvec support.

### See Also

- `?rvec::pmin` for argument types and recycling rules.
- [Calculate a minimum, maximum, or range](#min-max-range) for summaries
  across elements.
