---
title: "Cookbook for rvec Package"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cookbook for rvec Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

This cookbook collects recipes for common tasks with rvecs. Each recipe
starts with a problem, gives a solution, and explains when and how to use it.

```{r libs}
library(rvec)
```

# Summarise elements within draws

## Calculate a minimum, maximum, or range {#min-max-range}

### Problem

You have draws for rates in several regions. You want draws for the lowest
rate, highest rate, or range of rates across the regions.

### Solution

Use `min()`, `max()`, or `range()`. Each function calculates its result
independently within each draw.

Here each row is a region and each column is a draw. The region with the
lowest or highest rate can differ between draws.

```{r regional-rates}
m <- rbind(
  north = c(3, 1, 8, 4),
  south = c(1, 7, 3, 6),
  west = c(2, 4, 5, 2)
)
m
rate <- rvec(m)
```

```{r extrema}
lowest <- min(rate)
highest <- max(rate)
limits <- range(rate)

as.matrix(lowest)
as.matrix(highest)
as.matrix(limits)
```

`lowest` and `highest` are rvecs of length one, each retaining all four draws.
`limits` is an rvec of length two: its first element contains the minima and
its second contains the maxima. `range()` returns these two endpoints;
use `max(rate) - min(rate)` for the distance between them.

You can then summarise the draws of a derived quantity, just as with any
other rvec:

```{r summarise-highest}
draws_mean(highest)
```

### Discussion

`min(rate)` and `draws_min(rate)` answer different questions. `min(rate)`
finds the lowest regional rate within each draw. `draws_min(rate)` finds
the lowest draw for each region, returning an ordinary vector with one
value per region.

```{r compare-directions}
draws_min(rate)
```

Use `na.rm = TRUE` to remove missing values before calculating extrema.
For `range()`, `finite = TRUE` excludes missing values and both infinities.

You can supply additional rvecs or ordinary vectors, but put an rvec first.
An ordinary value is included in every draw:

```{r include-reference}
min(rate, 0)
```

### See Also

- `?min`, `?max`, and `?range` for base R's treatment of missing and empty inputs.
- `?draws_min` for minima and maxima across draws.
- [Troubleshooting: put the rvec first in `min()`](troubleshooting.html#min-argument-order).
