---
title: "Troubleshooting rvec"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Troubleshooting rvec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

This vignette collects solutions to problems encountered when using rvecs.
Each recipe shows a symptom, explains its cause, and demonstrates a fix.

```{r libs}
library(rvec)
```

# Summarise elements within draws

## Put the rvec first in min() {#min-argument-order}

### Problem

Calling `min()` with a numeric first argument and an rvec second argument
produces an error, even though both arguments contain numbers.

```{r min-numeric-first, error = TRUE}
x <- rvec(rbind(a = c(3, -1, 8), b = c(1, 7, -2)))
min(0, x)
```

### Solution

Put the rvec first:

```{r min-rvec-first}
answer <- min(x, 0)
as.matrix(answer)
```

The result has draws `0`, `-1`, and `-2`: the minimum of zero and all
elements of `x`, calculated separately within each draw.

### Discussion

Base R uses the first argument to decide whether to call the rvec method
for `min()`. With `min(0, x)`, the first argument is an ordinary number,
so R does not select that method. It then encounters the rvec's internal
list representation and reports an invalid-type error.

With `min(x, 0)`, R selects the rvec method, which aligns the arguments
and calculates a minimum within each draw. The same first-argument rule
applies to `max()` and `range()`.

```{r other-extrema}
as.matrix(max(x, 0))
as.matrix(range(x, 0))
```

Changing the argument order is enough; there is no need to convert the
rvec to an ordinary vector. Also note that `min(x, 0)` returns one element
per draw. It does not replace each positive element of `x` with zero.

### See Also

- [Cookbook: calculate a minimum, maximum, or range](cookbook.html#min-max-range).
- `?min`, `?max`, and `?range` for the base R functions.
