---
title: "vig1_overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vig1_overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Aims of **rvecs**

Bayesian analyses, stochastic simulations, and many other modern statistical methods produce random draws from distributions. Statistics calculated from these draws can be used to make inferences about the unknown quantities being modeled. For instance, medians across draws of can be used as point estimates for the unknown quantities. 

An enormously useful feature of random draws is that these draws can also be used to make inferences about *derived* quantities. The procedure is:

- Step 1. Calculate the derived quantity for each of the random draws.
- Step 2. Summarise the distribution of these derived quantities.

Say, for instance, that we had a set of draws for age-specific mortality rates, and we wanted to make inferences about life expectancy (a deterministic function of mortality rates). We would proceed as follows:

- Step 1. Derive life expectancy for each set of age-specific mortality rates
- Step 2. Calculate means, medians, or other statistics across these life expectancies.

Package **rvec** aims to make working with random draws fast and easy. **rvec** implements a data structure called an `rvec` that holds multiple random draws. An rvec can be transformed, coerced, or rearranged using essentially the same code that is used for ordinary vectors, which helps with Step 1 above. **rvec** contains functions for summarizing across random draws, for Step 2, that are sufficient for most purposes. For specialized summaries, and for visualising results derived distibutions, see packages such as [tidybayes](https://mjskay.github.io/tidybayes/),
[ggdist](https://mjskay.github.io/ggdist/), and [bayesplot](https://mc-stan.org/bayesplot/).

rvecs extend the `vctrs` class from package [vctrs](https://vctrs.r-lib.org), and inherits standard vector behaviors.  Many of the calculations are done internally using functions from [matrixStats](https://CRAN.R-project.org/package=matrixStats). Package [posterior](https://CRAN.R-project.org/package=posterior) provides alternative to rvecs for storing random draws.

# Examples

## Toy example

We begin with a toy example, to illustrate basic functionality.
We have 3 draws for a scalar unknown quantity, theta.

```{r}
library(rvec)
l <- list(c(3, 1, 0))
theta <- rvec(l)
theta
```

The header `<rvec_dbl<3>[1]>` tells us rvec `theta` holds
three random draws, each of which is a double
(see `base::double()`) and has length 1.

We can perform standard mathematical operations:
```{r}
exp(theta^2) + 1
```

`theta` recycles to match the length of
other vectors,

```{r}
theta + c(1, -1)
```

including other rvecs,


```{r}
beta <- rvec(list(c(-0.2, 0.4, 0.1),
                  c(-0.1, 0.1, 0.3)))
beta		  
```

```{r}
theta + beta
```

To summarise across random draws, we use
the `draws_*` functions, e.g.


```{r}
draws_mean(theta)
```


## Divorce rates in New Zealand

Our next example is more realistic. 
We will use some standard tidyverse packages

```{r}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
```
and work with a posterior sample
from a Bayesian model of divorce rates (per thousand)
in New Zealand, in 2020,

```{r}
divorce
```

First we convert the data from a 'data base' format,
where each row describes of a single draw, 
to an rvec format,
where multiple draws are folded into an rvec.

```{r}
divorce_rv <- divorce %>%
  collapse_to_rvec(value = rate)
divorce_rv
```

Now that the number of draws is large, means
and standard deviations across draws are displayed,
rather than the individual draws. (This
format is copied from `rvar`s in package [posterior](https://cran.r-project.org/web/packages/posterior/index.html).
`rvar`s are discussed in Section XXX.)

Define the 'total divorce rate' to be the
number of divorces a person could expect to
experience over their lifetime, given prevailing
divorces. The total divorce rate can be calculated
as

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(TDR = sum(rate) * 5 / 1000)
```

We can summarise across draws using function
`draws_quantile`. (`draws_quantile` returns a
tibble rather than a vector, so, following standard
`mutate` rules, we do not explicitly
create new columns.)

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(tdr = sum(rate) * 5 / 1000) %>%
  mutate(draws_quantile(tdr))
```

Next we look at the ratio between female and male divorce
rates

```{r}
divorce_rv %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(ratio = Female / Male)
```


```{r, fig.width = 7, fig.height = 4}
data <- divorce_rv %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(ratio = Female / Male) %>%
  mutate(draws_quantile(ratio))
ggplot(data,
       aes(x = age, 
           ymin = ratio_2.5, 
           y = ratio_50,
           ymax = ratio_97.5)) +
  geom_pointrange()
```



# Working with rvecs

## Creation

## Manipulation

## Mathematics

## Summarising
   
  
# Design

- vctrs
- matrixStats



# Other packages

## rv

## coda, posterior

## ggdist

?other









