---
title: "vig1_overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vig1_overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Aims of **rvecs**

Bayesian analyses, stochastic simulations, and other statistical methods produce random draws from distributions. Statistics calculated from these draws can be used to make inferences about the quantities being modeled. Medians calculated from draws can, for instance, be used as point estimates.

An enormously useful feature of random draws is that they can be used to make inferences about derived quantities. The procedure is:

- Step 1. Calculate the derived quantity for each of the random draws.
- Step 2. Summarise the distribution of these derived quantities.

If, for instance, we had a set of draws of age-specific mortality rates, and we wanted to make inferences about life expectancy (a deterministic function of mortality rates), then we could proceed as follows:

- Step 1. Derive life expectancy for each set of age-specific mortality rates.
- Step 2. Calculate means, medians, or other statistics for these life expectancies.

Package **rvec** aims to make working with random draws fast and easy. **rvec** provides a data structure called an rvec that holds multiple random draws. An rvec can be manipulated using essentially the same code that is used for ordinary vectors. **rvec** also contains functions for summarizing across random draws.

rvecs extend the `vctrs` class from package [vctrs](https://vctrs.r-lib.org), and inherits standard vector behaviors.  Many of the calculations are done internally using functions from [matrixStats](https://CRAN.R-project.org/package=matrixStats), for speed.

Package [posterior](https://CRAN.R-project.org/package=posterior) provides alternative data structures for holding random draws that preserve more information about the way the draws were generated, but that are less convenient to manipulate. Packages [tidybayes](https://mjskay.github.io/tidybayes/),
[ggdist](https://mjskay.github.io/ggdist/), and [bayesplot](https://mc-stan.org/bayesplot/) provide specialised ways of summarising draws, and tools for visualising distribtutions. Converting rvecs into and out of formats used by these packages is easy.

# Examples

## Toy example

We begin with a toy example, to illustrate basic functionality.

```{r}
library(rvec)
l <- list(c(3, 1, 0))
theta <- rvec(l)
theta
```

The header `<rvec_dbl<3>[1]>` tells us rvec `theta` holds three random draws, each of which is a double (see `base::double()`), and that, viewed as a vector, `theta` has length 1.

We can perform standard mathematical operations:
```{r}
exp(theta^2) + 1
```

`theta` recycles to match the length of other vectors,

```{r}
theta + c(1, -1)
```

including other rvecs,


```{r}
beta <- rvec(list(c(-0.2, 0.4, 0.1),
                  c(-0.1, 0.1, 0.3)))
beta		  
```

```{r}
theta + beta
```

To summarise across random draws, we use `draws_*` functions, e.g.

```{r}
draws_mean(theta)
```


## Divorce rates in New Zealand

Our next example is more involved, and includes the use of some standard tidyverse packages.

```{r}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
```

We analyse a posterior sample rom a Bayesian model of divorce rates
in New Zealand, by age and sex. The rates are divorces per thousand people (not necessarily married) per year.

```{r}
divorce
```

First we convert the data from a 'data base' format, where each row describes of a single draw, to an rvec format, where multiple draws are folded into an rvec.

```{r}
divorce_rv <- divorce %>%
  collapse_to_rvec(value = rate)
divorce_rv
```

Now that the number of draws is large, means and standard deviations across draws are displayed, rather than the individual draws. This display format is copied from the `rvar` data structure in package [posterior](https://cran.r-project.org/web/packages/posterior/index.html). rvars are discussed in Section XXX.

We define the 'total divorce rate' to be the number of divorces that a person could expect to experience over their lifetime, given prevailing
divorces. The total divorce rate can be calculated as

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(TDR = sum(rate) * 5 / 1000)
```

We can summarise across draws using function `draws_quantile`. `draws_quantile` returns a tibble rather than a vector, so, following standard `mutate` rules, we do not explicitly create new columns.

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(tdr = sum(rate) * 5 / 1000) %>%
  mutate(draws_quantile(tdr))
```

Next we calculate the ratio between female and male divorce rates,

```{r}
divorce_ratio <- divorce_rv %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(ratio = Female / Male) %>%
  mutate(draws_quantile(ratio))
```

and graph the result

```{r, fig.width = 7, fig.height = 4}
ggplot(divorce_ratio,
       aes(x = age, 
           ymin = ratio_2.5, 
           y = ratio_50,
           ymax = ratio_97.5)) +
  geom_pointrange()
```

# What are rvecs

- matrix
- functions applied in parallel
   - except for draw_* functions
- can be double, int, char, logical
- combine two: must have same draws


# Working with rvecs

## Creating



### Individual rvecs



### Whole data frames



## Manipulation

## Mathematics

## Summarising
   
  
# Design

- vctrs
- matrixStats



# Other packages

## rv

- early implementation, but defunctx

## rvar

- not just one extra dimension, can be matrix, array
- (with rvec, imagine it to be stretched out)
- 

## rv

- works inside data frames


## coda, posterior

## ggdist

?other









