---
title: "Overview of package rvecs"
output:
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    fig_caption: yes
    toc: true
    toc_depth: 2
    number_sections: true
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Overview of Package rvecs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Aims of **rvecs** package

Many modern statistical methods, including most Bayesian methods, produce random draws from a distribution. A exceptionally useful feature of these draws is that they can be used to make inferences about derived quantities. The procedure is:

- Step 1. Calculate the derived quantity for each of the random draws.
- Step 2. Summarise the distribution of these derived quantities.

If, for instance, we have randoms draws of age-specific mortality rates, and we want make inferences about life expectancy (a summary indicator for mortality rates), then we proceed as follows:

- Step 1. Derive life expectancy for each set of age-specific mortality rates.
- Step 2. Calculate means, medians, or other statistics for these life expectancies.

Package **rvec** provides tools for working with random draws. The draws are held in a structure called an rvec, which can, for many purposes, be treated like an ordinary R vector, and manipulated using ordinary base R and [tidyverse](https://www.tidyverse.org) code. **rvec** also contains functions for summarizing across random draws.




# Examples

## Toy example

We begin with a toy example, to illustrate basic functionality.

```{r}
library(rvec)
l <- list(c(3, 1, 0))
theta <- rvec(l)
theta
```

The header `<rvec_dbl<3>[1]>` indcates that rvec `theta` holds three random draws, each of which is a double, and that, regarded as a vector, `theta` has length 1.

We can perform standard mathematical operations:
```{r}
theta^2 + 1
```

`theta` recycles to match the length of other vectors,

```{r}
theta + c(1, -1)
```

including other rvecs,


```{r}
beta <- rvec(list(c(-0.2, 0.4, 0.1),
                  c(-0.1, 0.1, 0.6)))
beta		  
```

```{r}
theta + beta
```

To summarise across random draws, we use `draws_*` functions, e.g.

```{r}
draws_mean(beta)
```


## Divorce rates {#sec:divorce}

Our next example is more involved, and includes the use of some standard tidyverse packages.

```{r}
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(ggplot2)
```

We analyse a posterior sample from a Bayesian model of divorce rates in New Zealand. The rates are divorces per thousand people per year, disaggregated by age and sex.

```{r}
divorce
```

First we convert the data from a 'data base' format, where each row describes a single draw, to an rvec format, where multiple draws are folded into an rvec.

```{r}
divorce_rv <- divorce %>%
  collapse_to_rvec(value = rate)
divorce_rv
```

When the number of draws is large, the print method displays `<median> (<2.5% quantile>, <97.5% quantile>)` for the distribution, rather than the individual draws. 

We define the 'total divorce rate' to be the number of divorces that a person would expect to experience over their lifetime under prevailing divorce rates. The total divorce rate can be calculated as

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(TDR = sum(rate) * 5 / 1000)
```

We summarise across draws using `draws_ci()`, which, by default, calculates medians and 95% credible intervals. Function `draws_ci()` returns a tibble rather than a vector, so, following standard `mutate` rules, we do not explicitly create new columns.

```{r}
divorce_rv %>%
  group_by(sex) %>%
  summarise(tdr = sum(rate) * 5 / 1000) %>%
  mutate(draws_ci(tdr))
```

Next we calculate the ratio between female and male divorce rates,

```{r}
divorce_ratio <- divorce_rv %>%
  pivot_wider(names_from = sex, values_from = rate) %>%
  mutate(ratio = Female / Male) %>%
  mutate(draws_ci(ratio))
```

and graph the result

```{r, fig.width = 7, fig.height = 4}
ggplot(divorce_ratio,
       aes(x = age, 
           ymin = ratio.lower, 
           y = ratio.mid,
           ymax = ratio.upper)) +
  geom_pointrange()
```

# Structure of rvecs

The class `"rvec"` has four subclasses:

- `"rvec_dbl"`, which holds doubles, eg `3.142`, `-1.01`.
- `"rvec_int"`, which holds integers, eg `42`, `-1`.
- `"rvec_lgl"`, which holds `TRUE`, `FALSE`, and `NA`.
- `"rvec_chr"`, which hold characters, eg `"a"`, `"Thomas Bayes"`.

Internally, an rvec is a matrix, with each row representing one unknown quantity, and each column representing one draw from the joint distribution of the unknown quantities,

|              | Draw 1        | Draw 2        | $\dots$  | Draw $n$      |
|--------------|:-------------:|:-------------:|:--------:|:-------------:|
| Quantity 1   | $\theta_{11}$ | $\theta_{12}$ | $\dots$  | $\theta_{1n}$ |
| Quantity 2   | $\theta_{21}$ | $\theta_{22}$ | $\dots$  | $\theta_{2n}$ |
| $\vdots$     | $\vdots$      | $\vdots$      | $\ddots$ | $\vdots$      |
| Quantity $m$ | $\theta_{m1}$ | $\theta_{m2}$ | $\dots$  | $\theta_{mn}$ |


Ordinary functions are applied independently to each column. For instance, calling `sum()` on an rvec creates a new rvec with structure

|              | Draw 1                    | Draw 2                    | $\dots$  | Draw $n$                  |
|--------------|:-------------------------:|:-------------------------:|:--------:|:-------------------------:|
| Quantity 1   | $\sum_{i=1}^m\theta_{i1}$ | $\sum_{i=1}^m\theta_{i2}$ | $\dots$  | $\sum_{i=1}^m\theta_{in}$ |


Functions with a `draws_` prefix are applied independently to each row. For instance, calling `draws_mean()` on an rvec creates a new numeric vector with structure


|              | Value                                |
|--------------|:------------------------------------:|
| Quantity 1   | $\frac{1}{n}\sum_{j=1}^n\theta_{1j}$ |
| Quantity 2   | $\frac{1}{n}\sum_{j=1}^n\theta_{2j}$ |
| $\vdots$     | $\vdots$                             |
| Quantity $m$ | $\frac{1}{n}\sum_{j=1}^n\theta_{mj}$ |


Each rvec holds a fixed number of draws. Two rvecs can only be used together in a function if

1. both rvecs have the same number of draws, or
2. one or more rvecs have a single draw.



# Creating rvecs

An individual rvec can be created from

a list of vectors,
```{r}
x <- list(LETTERS, letters)
rvec(x)
```

a matrix, 
```{r}
x <- matrix(rnorm(2000), nrow = 2)
rvec(x)
```

or an atomic vector
```{r}
x <- c(TRUE, FALSE)
rvec(x)
```

Function `rvec()` chooses from classes `rvec_dbl`, `rvec_int`, `rvec_lgl`, and `rvec_chr`, based on the input. To enforce a particular class, use function `rvec_dbl()`, `rvec_int()`, `rvec_lgl()`, or `rvec_chr()`,

```{r}
x <- list(1:3)
rvec(x)
rvec_dbl(x)
rvec_chr(x)
```
When the raw data take the form of a database with one draw per row, the most efficient way to create rvecs is to use `collapse_to_rvec()`. See Section \@ref(sec:divorce) for an example.

# Mathematics on rvecs


# Probability distributions and rvecs



# Manipulating rvecs

## Subsetting

## Combining

## Coercing

## If-Else



# Summarising rvecs
   
  
# Design choices

rvecs extend the `vctrs` class from package [vctrs](https://vctrs.r-lib.org), and inherits standard vector behaviors.  Many of the calculations are done internally using fast functions from [matrixStats](https://CRAN.R-project.org/package=matrixStats).

- vctrs
- matrixStats



# Other packages

Packages [rv](https://CRAN.R-project.org/package=rv ) and [posterior](https://CRAN.R-project.org/package=posterior) contain similar data structures to rvecs, and package **rvec** borrows ideas from both. The advantages of rvecs is that they behave more like standard R vectors, which makes Step 1, the manipulation of draws, easier. 

Packages [tidybayes](https://mjskay.github.io/tidybayes/),
[ggdist](https://mjskay.github.io/ggdist/), and [bayesplot](https://mc-stan.org/bayesplot/) provide advanced tools for summarising draws, and for visualising distributions. Package **posterior** contains tools for evaluating the procedures that produced the random draws. **rvec** includes functions for converting into and out of the formats used by these packages, so that the packages can be used in combination.



## Functionality similar to **rvec**

### rv

- excellent paper on theory

### posterior

- not just one extra dimension, can be matrix, array
- (with rvec, imagine it to be stretched out)


## Working with draws

## coda, posterior

## ggdist

## tidybayes

?other









